---
title: "xxx"
author: 
  - Daniel Du
  - Michelle Ji
thanks: "Code and data are available at: [https://github.com/danield424/Ground_Beef_Price_Analysis/](https://github.com/danield424/Ground_Beef_Price_Analysis/)."
date: today
date-format: long
abstract: xxx
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
```


# Introduction
xxx


The music industry continues to increase significance in everyday life, from listening to music on a daily commute to school or work to music played in stores. Culture and region also plays a role in what artists and songs are popular, as seen with various Billboard 100 lists from different countries. This paper examines the difference between music listening in the United States and the United Kingdom, focusing on danceability as the main component for music listening habits. Our findings highlight key features of music that artists could use to see what is popular with their listeners the most, depending if they are a more popular artist in a specific region. 

The paper is broken down into various sections. @sec-data explores the data, looking at ground beef prices of different stores over time. @sec-results introduces observations and potential conclusions from the data analysis.


# Data {#sec-data}

We use @citeR and @tidyverse. We pull @spotify artist audio features data using @spotifyr a wrapper around Spotify’s API. We find the top three US artists from @billboard and the top three artists in the UK from @musicmetricsvault. We drop Chappell Roan as a top three candidate because of her short album history. The artists’ danceability are averaged out to create country-level danceability averages over time.

```{r, include=FALSE}
# Read in data 
#| echo: false

```



```{r}
#| echo: false
#| label: fig-1
#| warning: false
#| 

```







```{r}
#| echo: false
#| label: fig-2
#| warning: false

```

```{r}
#| echo: false
#| label: fig-3
#| warning: false


```




\newpage

# Results {#sec-results}

We find that the top three artists from the United Kingdom, which include Dua Lipa, Ed Sheeran, and Adele, tend to have higher danceability levels than the top three artists from the United States, which include Sabrina Carpenter, Taylor Swift, and Billie Eilish. @fig-1 details the average danceability across the top three United States artists. @fig-2 shows the average danceability across the top three United Kingdom artists. @fig-3 highlights the average danceability of top three artists in both the US and the UK. From this graph, we can see that the UK top three artists have higher average danceability levels than the US. 


\newpage


# References


